---
title: "Gene regulatory network inference with TIGRESS"
author: "Jean-Philippe Vert"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.show='hold', fig.width=3.4, fig.height=3.4)
set.seed(4395)
```


## Introduction

`tigress` implements the TIGRESS model of Haury et al. (2012) to infer a gene regulatory network from gene expression data. The input of `tigress` is a matrix of gene expression data, and a list of known transcription factors among the genes. The output is a scoring of all potential regulations from a transcription factor to any gene. To obtain a gene regulatory network, one can just keep regulations with a score above a user-defined threshold. In this vignette we illustrate the use of `tigress` to reconstruct the gene regulatory network of the E. coli bacteria. Before we start, we need to load the `tigress` package:
```{r}
library(tigress)
```

## Data

The `tigress` package comes with a gene expression dataset for E. coli. Each row is a gene, each column an experiment.
```{r}
dim(ecoli$exp)
ecoli$exp[1:5,1:3]
```
